---
title: "Merging Datasets Using R"
author: "Assil NOUREDDINE"
email: "noureddine.assil@stud.hs-fresenius.de"
matriculation-number: "400924750"
date: "11 June 2025"
format:
  html: default
  pdf: default
bibliography: references.bib
csl: apa.csl
---

# Welcome

## Merging Datasets Using R  
👩‍💻 Assil NOUREDDINE  
📧 noureddine.assil@stud.hs-fresenius.de  
📅 11 June 2025

---

# Agenda

1. Why merge datasets?  
2. Join types in R  
3. Tidyverse tools  
4. Hands-on examples  
5. Best practices  
6. Summary & References

---

# Why Merge Datasets?

- Data is often stored in multiple tables  
- Combining them gives more insights  
- Example: Employee data + Salary data [@wickham2023dplyr]

---

# Case Study

### Example CSVs:

**employees.csv**

| employee_id | name    | department |
|-------------|---------|------------|
| 1           | Alice   | HR         |
| 2           | Bob     | IT         |
| 3           | Charlie | Finance    |

**salaries.csv**

| employee_id | salary |
|-------------|--------|
| 2           | 60000  |
| 3           | 55000  |

---

# Join Types in R

| Join        | Description                        |
|-------------|------------------------------------|
| `inner`     | Only matching keys in both tables  |
| `left`      | All rows from left + matches       |
| `full`      | All rows from both                 |
| `anti`      | Only rows from left, not in right  |

[@wickham2023dplyr]

---

# Load the Tools

```{r}
library(tidyverse)

employees <- read_csv("Data/employees.csv")
salaries <- read_csv("Data/salaries.csv")

```
---
# Example 

inner_join(): 

```{r}
inner_join(employees, salaries, by = "employee_id")
```

Only includes matches
--- 

left_join():

```{r}
left_join(employees, salaries, by = "employee_id")
```

Includes all employees
Adds salary where available

--- 

anti_join():
```{r}
anti_join(employees, salaries, by = "employee_id")
```
Helps find records in employees not matched in salaries

---

# Visualizing with ggplot2

```{r}
left_merged <- left_join(employees, salaries, by = "employee_id")

ggplot(left_merged, aes(x = department, y = salary)) +
  geom_col(na.rm = TRUE) +
  labs(title = "Salaries by Department")
```
---

# Common Issues

- Duplicated keys — use distinct()
- A values — check with anti_join()
- Column type mismatch — fix with as.character() or as.numeric()
- Performance — consider data.table for large data [@rteam2025base]

---

# Best Practices

- Clean your data first
- Always check join output
- Document your steps
- Visualize where possible
- Cite your packages and sources [@apa2020manual]

---

# Summary

- Joins help combine useful data
- Different joins serve different purposes
- Tidyverse makes merging easier
- Joins + ggplot = insights!

---

# References



